Integration of national demographic-disturbance relationships and local data can improve caribou population viability projections and inform monitoring decisions

### 2025-02-21 analysis/paper/paper.Rmd
preparing EI submission

---
title: "Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country. Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and to reduce the risk that lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.  We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years may be better than intensive short-term monitoring for improving the reliability of status projections. 
keywords: |
  woodland caribou (*Rangifer tarandus caribou*); anthropogenic disturbance scenarios; population viability; value of information; monitoring decisions; species at risk; prior information
highlights: |

@@

  * We show how Bayesian integration of prior information from a national analysis of demographic-disturbance relationships can reduce uncertainty about the viability of sparsely monitored populations.

@@

  * At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years is likely better than intensive short-term monitoring for reducing uncertainty about population viability.

  * We introduce a method for acknowledging potential errors in recruitment surveys to help clarify where and when these these errors are likely to alter conclusions about population viability.

@@

The number of species threatened with extinction is high and getting
higher [@ipbes_global_2019]. Reliable information on the status of
populations and likely impacts of management are needed to inform
conservation action, but effective monitoring is challenging and costly
[@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;
@dunham_strategic_2023]. Resources to support conservation and recovery
are limited [@buxton_balancing_2022; @mccarthy_financial_2012;
@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk
[@martin_acting_2012], and some populations have been monitored to
extinction [@lindenmayer_counting_2013]. A recent analysis of action
plans for Canadian species at risk found that half of tasks are research and
monitoring, and this proportion is highest for species with higher risk
of extinction, highlighting a need to more carefully consider when,
where, and why additional information is needed to inform conservation
action [@buxton_balancing_2022]. Analysis of the implications of
monitoring outcomes for decisions can help to clarify whether additional
monitoring is necessary to inform conservation action
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012].

Analysis methods that can make better use of available monitoring
data to inform decisions include Bayesian population modelling
[@schaub_integrated_2021] and value of information analyses
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012]. For sparsely monitored populations, incorporating prior information from elsewhere can help reduce uncertainty [@drummond_combining_2018; @parlato_incorporating_2021; @tufto_bayesian_2000]. To be useful, these methods must be tailored
to particular species and decision contexts. Here, we offer a simple
Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates prior information about the impacts of landscape 
change [@johnson_science_2020] with available demographic data [building on @eacker_web-based_2019] to reduce uncertainty in
population viability projections. Following the value of information analysis
framework proposed by Dunham et al. (2023), we combine this Bayesian
model with simulations of plausible population trajectories
[@stewart_climate-informed_2023; @dyson_existing_2022] to (i) show how the
value of local monitoring varies with landscape condition, and (ii) assess
the ability of alternative monitoring scenarios to reduce the risk of
errors in population viability projections. Our approach is tailored to
boreal caribou, but could be adapted for other species at risk found in
multiple populations that are costly to monitor and sensitive to habitat
disturbance.

Boreal caribou are widely and sparsely distributed through the Canadian
boreal region across a gradient of anthropogenic disturbance
[@eccc_scientific_2011; @johnson_science_2020] that includes some of the
most intact and inaccessible places remaining on earth
[@hirsh-pearson_canadas_2022; @ibisch_global_2016;
@venter_sixteen_2016]. Across 58 boreal caribou study areas, both
survival and recruitment decrease with the percentage of the study area
that is disturbed, and no populations of caribou inhabiting areas with high anthropogenic disturbance are considered self-sustaining [@eccc_scientific_2011; @johnson_science_2020]. The mechanisms behind these
demographic-disturbance relationships include direct impacts on habitat
availability and use, and indirect impacts on predators and alternate
prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreases in
recruitment as disturbance increases are also observed over time in
places where there are historic demographic data
[@rudolph_demographic_2017; @hervieux_widespread_2013], but many places
lack sufficient data to characterize local demographic-disturbance
relationships [@johnson_science_2020]. There is also variation in
demographic rates among study areas, particularly where anthropogenic
disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
@fortin_forest_2017; @gagne_impact_2016], and this variation leads to
uncertainty in projections derived from national disturbance-demographic
relationships [@stewart_climate-informed_2023; @dyson_existing_2022].

Interventions such as wolf culls, maternal penning, and restoration of
linear features are intended to improve survival and reproduction of
populations that are not otherwise self-sustaining
[@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
need to evaluate and compare effects of boreal caribou conservation
actions with careful monitoring. Our focus in this paper is on assessing
and projecting self-sustainability of populations in the absence of
these interventions, to clarify when additional information is needed to
reduce uncertainty about whether there is a need for action. Where
actions are taken, outcomes should be monitored; the model we
propose here does not include effects of management interventions, and
cannot be used to project outcomes for managed populations.

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g. @flasko_fecal_2017;
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022; @eacker_web-based_2019; @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].
Various efforts have been made to improve caribou population viability
projections and growth rate estimates using Bayesian methods
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. An advantage of Bayesian demographic models is that they account for multiple sources of uncertainty
[@schaub_integrated_2011; @eacker_assessing_2017;
@schaub_integrated_2021]. However, there is very little monitoring data
available for some caribou populations [@johnson_science_2020] and there
is a risk that lack of local data will be used as a reason to delay
conservation action [@bissonnette_controverses_2020; @boan_climate_2018]. Bayesian caribou population
models [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that uncertainty in models
informed only by local demographic data could be used to delay conservation action for poorly monitored populations. We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships [@johnson_science_2020] with
available local demographic data in a simple Bayesian population model
for boreal caribou.

While we believe that using available information to reduce uncertainty
is important, we also believe it is important to acknowledge remaining
uncertainty. In general, parametric Bayesian population models include a
number of simplifying assumptions, and misspecified models can yield
biased or overconfident results [@schaub_integrated_2021]. More
specifically, models that do not account for observation errors can be
misleading [@schaub_integrated_2021]. Although several authors have
pointed out that misclassification of adults and missing calves in
composition surveys can bias recruitment estimates from those surveys
[@ellington_divergent_2020; @decesare_estimating_2012], none have thus
far examined sensitivity of Bayesian caribou population models to these
sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias term derived from a simple model of the recruitment survey observation
process.

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, can improve viability projections on changing
landscapes, and can help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach will help to identify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations [@bissonnette_controverses_2020; @boan_climate_2018].

@@

To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort we analyze simulated data using a Bayesian model with informative priors and a composition survey bias term (section 2.1). Steps in our analysis procedure include defining anthopogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4). Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, and the expected value of local information (section 2.5). We refer to the model used to simulate example trajectories as the "national" model because parameters are from a national analysis of demographic-disturbance relationships [@johnson_science_2020]. We distinguish this national model from the Bayesian model that is used to analyze simulated example demographic trajectories. Where appropriate the national model and the Bayesian model share the same structure, and we have selected notation to highlight the similarities. Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without interannual variation). For reference, all variables and parameters are defined in supplemental Table A1.

@@

resultScenario = "S14"
resultSet ="cmult6ay1aSf1"
figSet = c("diffslow","distScns","EVSI","power","EVPIall","powerLam","probViability")

@@

We focus on analyzing survival of radio-collared females and the ratio of calves to cows in aerial
surveys of groups that include collared females [@nbckc_nbckc_2022],
because these remain the most widely available sources of demographic
information for boreal caribou populations [@johnson_science_2020]. GPS radio collars report animal locations at regular intervals, and cease moving when animals die. Aerial recruitment surveys are usually done in March or April, and radio-collared females are often used to locate groups of animals. Surveyors use visual cues to distinguish calves that have survived their first winter (~ 9 months old) from adult cows and bulls. Surveyed groups also contain uncollared adults so sample size is typically larger for recruitment than for survival estimates. Errors that can bias composition survey results include misidentifying young
bulls as cows or vice versa, and failing to detect calves
[@ellington_divergent_2020; @decesare_estimating_2012].

To derive a composition survey bias term we assume each group
of animals in a calf:cow composition survey contains one or more
collared adult females ($T$), and may also include: uncollared adult
females misidentified as young bulls or unknown sex ($U$); correctly
identified uncollared adult females ($V$); young bulls correctly
identified as male or unknown sex ($O$); young bulls misidentified as
uncollared adult females ($P$); observed calves ($J$); and unobserved
calves ($K$). The apparent number of adult females in the group is
$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
adult females to collared animals in the recruitment survey.
The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
$P=(O+P)u$. Given a probability $z$ of missing calves, we get
$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
bias-adjusted recruitment rate $X$ can be written as a function of the
observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
bulls to uncollared adult females $q$, and the misidentification probabilities $u$
and $z$ (see Supplement B for details). For simplicity, we combine the
probabilities into a bias term $c$ that is integrated into
the Bayesian population model and simulations (below):

@@

To investigate the effectiveness of various monitoring strategies, and to show how this
varies with landscape condition, we begin by simulating example demographic trajectories and observations of those trajectories. To simulated example trajectories we use a modified version of Johnson et al's [-@johnson_science_2020] demographic model described by Dyson et al. [-@dyson_existing_2022]. For brevity, we refer to this as the "national" model, and for clarity we include a full description in the following paragraphs. Together, a set of example trajectories from the national model represent alternative hypotheses about the true state of a particular caribou population. To assess the accuracy of population status assessments and the expected value of alternative monitoring strategies [@runge_which_2011; @dunham_strategic_2023] we analyze each trajectory of simulated observations using a Bayesian model (section 2.4). We also use the sets of simulated trajectories to confirm that prior predictions from our Bayesian model are consistent with knowledge of the observed distribution of outcomes for caribou across the country [@johnson_science_2020].

@@

Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
population sizes, and declines to $0.6\dot{X}_t$ as the population
reaches carrying capacity of $k=10000$.
The post-juvenile female population in the next year includes both
survivors and new recruits:

\begin{equation}
\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t),
  (\#eq:dotN)
\end{equation}

where the maximum potential population growth rate is $r_{max}=1.3$.

@@

In our simulated example populations, mortality risk does not vary over
among seasons or individuals, so the number of collared cows at the time of the composition survey $\hat{T}_t$ depends on the true survival rate $\dot{S}_t$ and the number of collared cows at the beginning of the year $\hat{I}_t$ as follows:
\begin{equation}
\hat{T}_t \sim \text{Binomial}(\hat{I}_t,\dot{S}_t).
  (\#eq:hatI)
\end{equation}

The number of cows in the composition survey $\hat{W}_t$ is given by the
number of collared cows $\hat{T}_t$ and the apparent number of adult
females per collared female observed in the composition survey $w$:
\begin{equation}
\hat{W}_t = w\hat{T}_t.
  (\#eq:hatW)
\end{equation}
The number of observed calves $\hat{J}_t$ also depends on
the unadjusted apparent recruitment for the population $\dot{R}_t$:
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,\dot{R}_t).
  (\#eq:hatJ)
\end{equation}

@@

Where appropriate, the Bayesian and national population models share the same structure, and we have selected notation to highlight the similarities. The difference between equations
\@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an
unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to
simulate observations. In the Bayesian model, in contrast, observations
are provided, and used to estimate the unobserved recruitment rate
$R_t$.

@@

Survival $S_t$ is estimated from the observed number of collared cows at the beginning of each year $\hat{I}_t$ and the number of those that survive the year $\hat{T}_t$ assuming a Binomial distribution:
\begin{equation}
\hat{T}_t \sim \text{Binomial}(\hat{I}_t,S_t).
  (\#eq:hatSEst)
\end{equation}

>>>>

---
title: "Integration of national demographic-disturbance relationships and local data can improve caribou population viability projections and inform monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but no populations inhabiting areas with high anthropogenic disturbance are considered self-sustaining. Demographic projections derived from local population-specific data are uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to improve population viability projections, and to reduce the risk that a lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves, through a term derived from a simple model of the recruitment survey observation process. We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where initial uncertainty is high, each additional year of monitoring provides less new information. The estimated probability of viability indicates whether more information is needed to improve accuracy of population viability projections. 
keywords: |
  woodland caribou (*Rangifer tarandus caribou*); anthropogenic disturbance scenarios; population viability; value of information; monitoring decisions; species at risk; prior information
highlights: |

@@

  * We show how Bayesian integration of local data and prior information from a national analysis of demographic-disturbance relationships can reduce uncertainty about the viability of sparsely monitored populations.

@@

  * At lower disturbance levels where uncertainty is higher there are diminishing returns on additional monitoring, and the estimated probability of viability indicates whether more information is needed.

  * We introduce a method for acknowledging potential errors in recruitment surveys to help clarify where and when these errors are likely to alter conclusions about population viability.

@@

The number of species threatened with extinction is high and getting higher [@ipbes_global_2019]. Reliable information on the status of populations and likely impacts of management are needed to inform conservation action, but effective monitoring is challenging and costly [@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;@dunham_strategic_2023]. Resources to support conservation and recovery are limited [@buxton_balancing_2022; @mccarthy_financial_2012;@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013]. A recent analysis of action plans for Canadian species at risk found that half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022]. Analysis of the implications of monitoring outcomes can help to clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;@gregory_structured_2012].

Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022; @ibisch_global_2016;
@venter_sixteen_2016]. Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the area that is disturbed, and no populations inhabiting areas with high anthropogenic disturbance are considered viable in the absence of interventions such as predator control or maternal penning (i.e. self-sustaining) [@eccc_scientific_2011; @johnson_science_2020]. Identifying populations that are declining is important for clarifying where and when conservation action is needed. Disturbance affects caribou demography both through direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreasing recruitment with increased disturbance has also been observed over time where historic demographic data exist [@rudolph_demographic_2017; @hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
@fortin_forest_2017; @gagne_impact_2016], which leads to uncertainty in projections that rely only on national-scale relationships [@stewart_climate-informed_2023; @dyson_existing_2022].

Analysis methods that can make better use of available monitoring data to inform conservation decisions include Bayesian population modelling [@schaub_integrated_2021] and value of information analyses [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;@gregory_structured_2012]. For sparsely monitored populations, incorporating prior information from elsewhere can help reduce uncertainty [@drummond_combining_2018; @parlato_incorporating_2021; @tufto_bayesian_2000]. Various efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021; @mcnay_demographic_2022;@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. An advantage of Bayesian demographic models is that they account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. However, Bayesian caribou population models developed to-date [@moeller_integrating_2021; @mcnay_demographic_2022;@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not included prior information about the observed distribution of demographic rates across the country that could reduce uncertainty about the status of poorly monitored populations.

Here, we offer a simple Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates prior information about the impacts of landscape change [i.e. demographic-disturbance relationships in @johnson_science_2020] with available demographic data to reduce uncertainty in population viability projections. Note that for the purposes of this analysis we consider a population to be viable if the population growth rate is > 0.99. We also offer a novel method of acknowledging and accounting for errors in composition surveys. Following a value information analysis framework for conservation status assessments [@dunham_strategic_2023], we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023; @dyson_existing_2022] to (i) show how the value of local monitoring varies with landscape condition, and (ii) assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population viability [@eccc_scientific_2011; @johnson_science_2020] because it accounts for uncertainties in both local data and demographic-disturbance relationships. We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and improve the accuracy of population viability projections on changing landscapes. This approach can help clarify where available information is – and is not – sufficient for assessing and projecting population viability, and thereby reduce the risk that uncertainty is used as a reason to delay conservation action for imperiled populations [@bissonnette_controverses_2020; @boan_climate_2018]. Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.

@@

To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort, we analyze simulated data using a Bayesian model with informative priors (section 2.4) and a composition survey bias term (section 2.1). Steps in our analysis procedure include defining anthopogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4). Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, the posterior probability of population viability, and the expected value of local information (section 2.5). We refer to the model used to simulate example trajectories as the "national" model because parameters are from a national analysis of demographic-disturbance relationships [@johnson_science_2020]. We distinguish this national model from the Bayesian model that is used to analyze simulated example demographic trajectories. Where appropriate the national model and the Bayesian model share the same structure, and we have selected notation to highlight the similarities. Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without stochastic interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without stochastic interannual variation). For reference, all variables and parameters are defined in supplemental Table A1.

@@

resultScenario = "S13"
resultSet ="cmult6ay1aSf1"
figSet = c("diffslow","distScns","EVSI","power","EVPIall","powerLam","probViability")

erSums = read.csv(here::here(paste0("tabs/",resultScenario,"/ErrorsByBandcmult6ay1aSf1.csv")))

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g. @flasko_fecal_2017;
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012]. Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].

While we believe that using available information to reduce uncertainty
is important, we also believe it is important to acknowledge remaining
uncertainty. In general, parametric Bayesian population models include a
number of simplifying assumptions, and misspecified models can yield
biased or overconfident results [@schaub_integrated_2021]. More
specifically, models that do not account for observation errors can be
misleading [@schaub_integrated_2021]. Although several authors have
pointed out that misclassification of adults and missing calves in
composition surveys can bias recruitment estimates from those surveys
[@ellington_divergent_2020; @decesare_estimating_2012], none have thus
far integrated these sources of error into Bayesian caribou population models [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias term derived from a simple model of the recruitment survey observation process.

To derive a composition survey bias term we assume each group
of animals in a calf:cow composition survey contains one or more
collared adult females ($T$), and may also include: uncollared adult
females misidentified as young bulls or unknown sex ($U$); correctly
identified uncollared adult females ($V$); young bulls correctly
identified as male or unknown sex ($O$); young bulls misidentified as
uncollared adult females ($P$); observed calves ($J$); and unobserved
calves ($K$). The apparent number of adult females in the group is
$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
adult females to collared animals in the recruitment survey.
The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
$P=(O+P)u$. Given a probability $z$ of missing calves, we get
$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
bias-adjusted recruitment rate $X$ can be written as a function of the
observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
bulls to uncollared adult females $q$, and the misidentification probabilities $u$
and $z$ (see Supplement B for details). For simplicity, we combine the
probabilities into a bias term $c$ that is integrated into the national simulation model (section 2.4) and
the Bayesian model (section 2.5):

@@

To investigate the effectiveness of various monitoring strategies, and to show how this
varies with landscape condition, we begin by simulating example demographic trajectories and observations of those trajectories using a modified version of Johnson et al's [-@johnson_science_2020] demographic model described by Dyson et al. [-@dyson_existing_2022]. For brevity, we refer to this as the "national" model, and for clarity we include a full description in the following paragraphs. Together, a set of example trajectories from the national model represent alternative hypotheses about the true state of a particular caribou population. To assess the accuracy of population status assessments and the expected value of alternative monitoring strategies [@runge_which_2011; @dunham_strategic_2023] we analyze each trajectory of simulated observations using the Bayesian model (section 2.4). We also use the sets of simulated trajectories to confirm that prior predictions from our Bayesian model are consistent with knowledge of the observed distribution of outcomes for caribou across the country [@johnson_science_2020].

@@

Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
population sizes, and declines to $0.6\dot{X}_t$ as the population
reaches carrying capacity of $k=10000$.
The post-juvenile female population in the next year includes both
survivors and new recruits: $\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t)$, where the maximum potential population growth rate is $r_{max}=1.3$.

@@

In our simulated example populations the number of collared cows at the time of the composition survey $\hat{T}_t$ depends on the true survival rate $\dot{S}_t$ and the number of collared cows at the beginning of the year $\hat{I}_t$: $\hat{T}_t \sim \text{Binomial}(\hat{I}_t,\dot{S}_t)$. The number of cows in the composition survey $\hat{W}_t$ is given by the number of collared cows $\hat{T}_t$ and the apparent number of adult
females per collared female observed in the composition survey $w$: $\hat{W}_t = w\hat{T}_t$. The number of observed calves $\hat{J}_t$ also depends on the unadjusted apparent recruitment for the population $\dot{R}_t$:

\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,\dot{R}_t).
  (\#eq:hatJ)
\end{equation}

@@

The difference between equations \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an
unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to
simulate observations. In the Bayesian model, in contrast, observations
are provided, and used to estimate the unobserved recruitment rate
$R_t$.

@@

Survival $S_t$ is estimated from the observed number of collared cows at the beginning of each year $\hat{I}_t$ and the number of those that survive the year $\hat{T}_t$ assuming a Binomial distribution: $\hat{T}_t \sim \text{Binomial}(\hat{I}_t,S_t)$. Note that real known-fate radio collar data should be analyzed with a method that accounts for variation in survival rate and the number of collared animals throughout the year [e.g. @dalgarno_bboutools_2024; @eacker_web-based_2019; @lamb_effectiveness_2024]. However, our implementation of a two-step non-parametric survival analysis method [following @eacker_web-based_2019] yielded slightly biased survival estimates.
